---
title: 'Day 2: Regress to Impress'
author: Amanda Devine
date: '2022-04-05'
slug: diamonds-regression
categories:
  - ''
tags:
  - 100doda
  - diamonds
  - R
  - linear regression
description: ''
thumbnail: ''
---

```{r}
library(ggplot2)
```


In order to properly fit a linear regression model, there are certain assumptions about the dataset that need to be met:
- the observations are independent and identically distributed (i.i.d.)
- the data are heteroscedastic (variance is the same throughout the set)
- the predictors and response are linearly related
- the errors are normally distributed

These assumptions can be checked by taking a look at the data:

```{r}
plot(diamonds$carat, diamonds$price)
```

It looks as though there is some heteroscedasticity 


# Fitting a linear regression model

In order to check our assumptions, we can fit an initial linear regression model and check the residuals.

```{r}
model1 <- lm(price ~ carat + cut + color + clarity, data = diamonds)
summary(model1)
```

```{r}
d1 <- data.frame(carat = 4, color = 'D', clarity = 'IF', cut = 'Ideal')
d1
```

```{r}
predict(model1, d1, interval = 'prediction')
```


```{r}
d2 <- data.frame(carat = 1.75, color = 'D', clarity = 'IF', cut = 'Ideal')
predict(model1, d2, interval = 'prediction')
```

```{r}
par(mfrow = c(2, 2))
plot(model1)
```

```{r}
diamonds$log_price <- log(diamonds$price)
head(diamonds)
```

```{r}
model2 <- lm(log_price ~ carat + cut + color + clarity, data = diamonds)
summary(model2)
```

```{r}
par(mfrow = c(2, 2))
plot(model2)
```

